Feasibility of Harvesting Power To Run A Domestic Water Meter Using Streaming Cell Technology

### 2014-10-03 paper.tex
Friday morning changes

\documentclass[10pt,final,journal]{IEEEtran}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{siunitx}
% \usepackage[dvipsnames]{color}

@@

    \begin{abstract}
        We investigate the possibility of using streaming cells as a means of harvesting energy from a town water supply.
        We verify streaming cell measurements found in the literature and extend those measurements over a wider range of dimensions.
        We measure the electrical power developed streaming cells using town water.
        We estimate the amount of energy required to operate a simple data logger and transmitter.
        We estimate the amount of energy available from a typical domestic household based on water usage data.
        From these estimates we calculate minimum geometric dimensions required to deliver practical power output to a wireless water meter.
        Finally, we comment on the feasibility of using streaming cell technology as a means of harvesting energy from a domestic water supply.
    \end{abstract}

    \section{Introduction}
    Domestic and commercial water metering is becoming increasingly common throughout the world.~\cite{Chang2012}
    Cheap and reliable methods of retrieving meter data is an important consideration for utility companies.
    The introduction of wireless automatic meter readers offers many advantages to suppliers of town water.
    These advantages include reductions in cost and error rates associated with collecting metering data, increased billing frequency, leak detection and the elimination of the need for meter readers to access the consumers' property.~\cite{Chang2012,Britton2013}

@@

    Streaming cells provide a way of converting fluid energy into electrical energy without moving parts.
    They work on the principle that charged surfaces in contact with ionic solutions attract counter-ions\footnote{The terms counter-ion and co-ion refer to ions of opposite and like charge relative to the polarity of the solid surface respectively.} so as to form a charged layer at the contact surface.~\cite{Stein2004}
    This charged layer is termed the double layer and is the key mechanism that enables streaming cell energy conversion.
    The conversion is made by forcing ions within the double layer through narrow channels with the application of hydrostatic pressure.

    A double layer is called such because it is made of two individual layers; the Stern and diffuse layers.
    The Stern layer is comprised of immobilised ions situated at the charged solid surface.~\cite{Salieb-Beugelaar2009}
    The diffuse layer surrounds the Stern layer where counter-ions are not so strongly bound as to be immobile.
    A streaming cell moves ions in the diffuse layer through one or more narrow channels, concentrating them on the other side.

    Streaming cell channels can be created individually using a range fabrication methods, such as chemical etching or the use of narrowly separated parallel plates.
    Channels can also be formed en-mass by using porous materials such as glass or ceramics as the cell where the pores themselves form channels.
    Glass has the attractive property that it obtains a negative surface charge when in contact with water.
    This surface charge is caused by the deprotonation of surface silanol groups in glass (SiOH~$\leftrightarrows$~SiO$^{-}+$~H$^{+}$).~\cite{Kirby2004a}
    By immersing a glass channel in an electrolyte solution, double layers of positive ions occupy the interior walls of the channel.
    The thickness of the double layer is defined as the Debye length ($l_{D}$) and is controlled the ionic concentration of the liquid.~\cite{Israelachvili2011}

    When the dimensions of a channel become small enough, double layers will overlap.
    Overlapping double layers mean that co-ions are repelled from within the channel.

@@

    Streaming currents have been heavily investigated as a means of generating electrical energy from pressure gradients.~\cite{Chang2009,Daiguji2006,Daiguji2004b,Davidson2008a,Davidson2008,CherngHon2012,Jiao2014,Lu2006,Olthuis2005,Osterle1964,Pennathur2007,Ren2008a,VanderHeyden2006,Heyden2007,Xie2008,Yang2003}
    Theoretical predictions of the efficiency of standard micro/nano-fluidic channels are 2\% for pure water and 7\% for sodium chloride.~\cite{VanderHeyden2006}
    Experimental results show conversion efficiencies in the range of:
    \begin{itemize}
        \item 0.01\% by forcing water through porous glass with pore sizes from 10\thinspace--\SI{16}{\micro\metre}.~\cite{Yang2003}
        \item 0.8\% by forcing pure water through a ceramic rod populated with \SI{6}{\micro\metre} pores.~\cite{Yang2004}
        \item 3\% by forcing sodium chloride solution through a \SI{75}{\nano\metre} by \SI{50}{\micro\metre} silica channel.~\cite{Heyden2007}
        \item 0.77\% by forcing sodium chloride through a \SI{200}{\nano\metre} pore in an alumina membrane.~\cite{Lu2006}
        \item 5\% by forcing sodium chloride solution through a \SI{0.5}{\nano\metre} cylindrical pore in polyethylene terephalate foil.~\cite{Xie2008}
    \end{itemize}
    These results indicate that solutions containing salt are more efficient as a working liquid, however the maximum efficiency is found when the salt concentration is low.~\cite{VanderHeyden2006}

    It is clear from the literature that there is significant progress to be made with respect to increasing the conversion efficiency.
    Techniques to induce hydrodynamic slip at the fluid-solid interface are predicted to increase this efficiency to 30-40\%.~\cite{Davidson2008a, Ren2008a}
    Experimental results utilising slip enhanced channels have not yet been reported in the literature.
    This type of enhancement would overcome the `no-slip boundary condition' at the solid-liquid interface.
    Slip at the interface boundary would mean that ions in the Stern layer, where ionic concentration is highest, could also be moved across the channel.
    Surface enhanced channels are out of the scope of this study due to cost and manufacturing difficulty.

    % In~\cite{Joly2006} it is suggested that a polarised hydrophobic surface would strongly enhance electro-osmotic effects and therefore further increase conversion efficiency

@@

    The remainder of this work is presented as follows.
    In section \ref{sect:streamingCell} we build and measure a simple streaming potential cell.
    We confirm the relationship between pressure and applied pressure and investigate the effect of channel height.
    In section \ref{sect:waterConsumption} we estimate the amount of harvestable energy from a constructed domestic water usage profile.
    Section \ref{sect:powerRequirements} quantifies energy requirements of a simple electronic meter.
    Following on this, section \ref{sect:harvesterSize} presents the required cell dimensions of a harvester capable of producing the requisite energy.
    In section \ref{sect:results} we combine and present results of our measurements and quantify our predictions.
    And finally, we conclude with a discussion on the feasibility of using streaming cells to harvest energy.

@@

    We replicate the approximate dimensions of \cite{Gu2000} to produce ten streaming cells (\SI{245}{\micro\meter}, \SI{178}{\micro\meter}, \SI{161}{\micro\meter}, \SI{125}{\micro\meter}, \SI{106}{\micro\meter}, \SI{75}{\micro\meter}, \SI{71}{\micro\meter}, \SI{56}{\micro\meter}, \SI{52}{\micro\meter}, \SI{26}{\micro\meter}).
    The channels are made from glass microscope slides (Sail Brand - JIA 7101WT) sectioned into halves.
    Plastic shims (Garlock Colorplast) are epoxied (Selleys Araldite Ultra Clear Resin) between slide halves to separate the slides.
    The dimensions at each end of each channel are measured under a microscope to determine the channel height once the channels had set.

@@

    Pressure is applied across each cell by connecting one end to a tap and leaving the other open to atmosphere.
    A Honeywell pressure sensor (model 26PC15SMT) is placed across the cell to measure applied pressure.
    An Agilent precision measurement mainframe (model E5270B) is used to measure streaming potential, streaming current and to read the pressure sensor's output.
    The working fluid, being tap water, had a conductivity of \SI{183e-6}{\siemens\per\meter} as measured with an EDT Instruments RE 388Tx Conductivity Meter.

    Measured pressure-to-voltage gradients from each of the ten streaming cells are presented in Fig.~\ref{fig:cellEfficiency}.
    Spread in the data suggests uncertainty in the internal dimensions of each cell.
    Due to the opacity of the resin used, measurement of the internal cell dimensions was not possible.
    The plot shows the relationship between channel height and voltage gradient per Pascal of applied pressure.
    Each cell had a different flow rate, which this plot does not take into account.
    Three of the cells bust before flow measurements were obtained.
    % The \SI{26}{\micro\meter} channel shows a much lower gradient due to . which we expect is a result of a reduced flow rate.
    % The \SI{26}{\micro\meter} cell burst before we obtained flow rate measurements.

@@

    \section{Estimation of Harvestable Energy From A Typical New Zealand Dwelling}
    \label{sect:waterConsumption}
    \begin{figure}
        \begin{center}
        \includegraphics[width=\linewidth]{graph_profile}
        \end{center}
        \caption{Sample profile showing constructed instances of washing machine use, a shower and a toilet flush.
        Washing machine usage is broken into two parts corresponding to a wash and rinse cycle.}
        \label{fig:profileSample}
    \end{figure}

@@

    % \section{Estimation of Streaming Cell Harvester}
    % \label{sect:harvesterSize}
    \section{Conclusion}
    We have calculated approximately \SI{1}{\kilo\joule} of energy is dissipated per day inside a typical water meter at a two occupant dwelling.
    We have estimated that a harvester would need to harvest approximately \SI{10}{\joule} of energy per day to record and transmit water consumption data.
    Therefore an energy harvester with an energy conversion efficiency of 1\% is required.
    Streaming cells we constructed produced conversion efficiencies in the order of $1\times 10^{-6}\%$.

>>>>

\documentclass[10pt,final,journal]{IEEEtran}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[dvipsnames]{color}

@@

    \begin{abstract}
        We investigate the possibility of using streaming cells as a means of harvesting energy from the town water supply.
        We measure the electrical power developed from streaming cells using tap water.
        We estimate the amount of energy available from a typical domestic household based on water usage data.
        We estimate the amount of energy required to operate a simple data logger and transmitter.
        From these estimates we calculate the required efficiency and physical form of a streaming cell energy converter.
        We comment on the feasibility of using streaming cell technology as a means of harvesting energy from a domestic water supply.
    \end{abstract}

    \section{Introduction}
    Domestic and commercial water metering is becoming increasingly common throughout the world.~\cite{Chang2012}
    Cheap and reliable methods of retrieving metered data is an important consideration for utility companies.

    The introduction of wireless automatic meter readers offers many advantages to suppliers of town water.
    These advantages include reduced error rates associated with collecting data, increased billing frequency, leak detection and the elimination of the need for meter readers to access the consumers' property.~\cite{Chang2012,Britton2013}

@@

    Streaming cells provide a way of converting fluid energy into electrical energy without moving parts.
    They work on the principle that charged surfaces in contact with ionic solutions attract counter-ions so as to form a charged layer at the contact surface.~\cite{Stein2004}
    This charged layer is termed the double layer and is the key mechanism that enables streaming cell energy conversion.
    The conversion is made by forcing ions within the double layer through narrow channels with the application of hydrostatic pressure.

    \begin{figure}
        \begin{center}
        \includegraphics{diagram_streamingCellPrinciple}
        \end{center}
        \caption{Conceptualised rendering of the cross section of a streaming cell.
        Positive ions are being forced through the cavity by the application of pressure.}
        \label{fig:streamingPrinciple}
    \end{figure}

    A double layer is called such because it is made of two individual layers; the Stern and diffuse layers.
    The Stern layer is comprised of immobilised ions at the charged solid surface.~\cite{Salieb-Beugelaar2009}
    The diffuse layer surrounds the Stern layer and contains counter-ions that are not so strongly bound as to be immobile.
    The thickness of the double layer is defined as the Debye length ($l_{D}$) and is dependant upon the ionic concentration of the liquid.~\cite{Israelachvili2011}
    When the dimensions of a channel become small enough, double layers will overlap.
    Overlapping double layers mean that any co-ions are repelled from within the channel.

    The channels can be created individually using a range fabrication methods, such as chemical etching or the use of narrowly separated parallel plates.
    Channels can also be formed en-mass by using porous materials such as glass or ceramics as the cell where the pores themselves form channels.
    Glass has the attractive property that it obtains a negative surface charge when in contact with water.
    This surface charge is caused by the deprotonation of surface silanol groups in glass (SiOH~$\leftrightarrows$~SiO$^{-}+$~H$^{+}$).~\cite{Kirby2004a}
    By immersing a glass channel in an electrolyte solution, double layers of positive ions (counter-ions in this case) occupy the interior walls of the channel.

@@

    Streaming currents have been heavily investigated as a means of generating electrical energy from pressure gradients.~\cite{Chang2009,Daiguji2006,Daiguji2004b,Davidson2008a,Davidson2008,CherngHon2012,Jiao2014,Lu2006,Olthuis2005,Osterle1964,Pennathur2007,Ren2008a,VanderHeyden2006,Heyden2007,Xie2008,Yang2003}
    Theoretical predictions of the efficiency of standard micro/nano-fluidic channels are 2\% for pure water and 7\% for sodium chloride.~\cite{VanderHeyden2006}
    Experimental results show conversion efficiencies in the range of:
    \begin{itemize}
        \item 0.01\% by forcing water through porous glass with pore sizes from 10\thinspace--\SI{16}{\micro\metre}.~\cite{Yang2003}
        \item 0.8\% by forcing pure water through a ceramic rod populated with \SI{6}{\micro\metre} pores.~\cite{Yang2004}
        \item 3\% by forcing sodium chloride solution through a \SI{75}{\nano\metre} by \SI{50}{\micro\metre} silica channel.~\cite{Heyden2007}
        \item 0.77\% by forcing sodium chloride through a \SI{200}{\nano\metre} pore in an alumina membrane.~\cite{Lu2006}
        \item 5\% by forcing sodium chloride solution through a \SI{0.5}{\nano\metre} cylindrical pore in polyethylene terephalate foil.~\cite{Xie2008}
    \end{itemize}
    These results indicate that small channels using solutions containing salt are more efficient.
    According to \cite{Daiguji2004b}, the efficiency is maximised when the channel height is about half the Debye length.
    For water with a sodium content in the range 1--100~ppm this equates to a channel height in the order of 100--10~nm.
    Additionally, ~\cite{VanderHeyden2006} states that the maximum efficiency is found when the salt concentration is low.

    It is clear from the literature that there is significant progress to be made with respect to increasing the conversion efficiency.
    Techniques to induce hydrodynamic slip at the fluid-solid interface are predicted to increase this efficiency to 30-40\%.~\cite{Davidson2008a, Ren2008a}
    Experimental results utilising slip enhanced channels have not yet been reported in the literature.
    This type of enhancement would overcome the `no-slip boundary condition' at the solid-liquid interface.
    Slip at the interface boundary would mean that ions in the Stern layer, where ionic concentration is highest, could also be moved through the channel.
    Surface enhanced channels are out of the scope of this study due to cost and manufacturing difficulty.

@@

    The remainder of this work is presented as follows.
    In section \ref{sect:streamingCell} we build, measure and calculate the conversion efficiency of a simple streaming potential cell.
    We confirm the relationship between pressure and applied pressure and investigate the effect of channel height.
    In section \ref{sect:waterConsumption} we estimate the amount of harvestable energy from a domestic water usage profile.
    Section \ref{sect:powerRequirements} quantifies energy requirements of a simple electronic meter.
    And finally, we conclude with a discussion on the feasibility of using streaming cells to harvest energy.

@@

    We replicate the approximate dimensions of \cite{Gu2000} to produce ten streaming cells.
    The approximate internal channel heights of these channels are \SI{245}{\micro\meter}, \SI{178}{\micro\meter}, \SI{161}{\micro\meter}, \SI{125}{\micro\meter}, \SI{106}{\micro\meter}, \SI{75}{\micro\meter}, \SI{71}{\micro\meter}, \SI{56}{\micro\meter}, \SI{52}{\micro\meter}, and \SI{26}{\micro\meter}.
    The channels are made from glass microscope slides (Sail Brand - JIA 7101WT) sectioned into halves.
    Plastic shims (Garlock Colorplast) are epoxied (Selleys Araldite Ultra Clear Resin) between slide halves to separate the slides.
    The dimensions at each end of each channel are measured under a microscope to determine the channel height once the channels had set.

@@

    Pressure is applied to the cell by connection to tap and leaving the output connection open to atmosphere.
    A Honeywell pressure sensor (model 26PC15SMT) is placed across the cell to measure applied pressure.
    An Agilent precision measurement mainframe (model E5270B) is used to measure streaming potential, streaming current and to read the pressure sensor's output.
    The working fluid, being tap water, had a conductivity of \SI{183e-6}{\siemens\per\meter} as measured with an EDT Instruments RE 388Tx Conductivity Meter.

    Measured pressure-to-voltage gradients from each of the ten streaming cells are presented in Fig.~\ref{fig:cellEfficiency}.
    Spread in the data is attributed to uncertainty in the internal dimensions of each cell.
    Due to the opacity of the resin used, measurement of the internal cell dimensions was not possible.
    The plot shows the relationship between channel height and voltage gradient per Pascal of applied pressure.
    Each cell had a different flow rate, which this plot does not take into account.
    Three of the cells burst before flow measurements were obtained.
    % The \SI{26}{\micro\meter} channel shows a much lower gradient due to . which we expect is a result of a reduced flow rate.
    % The \SI{26}{\micro\meter} cell burst before we obtained flow rate measurements.

@@

    \section{Estimation of Harvestable Energy From A Typical New Zealand Dwelling}
    \label{sect:waterConsumption}

    % \begin{figure}
    %     \begin{center}
    %     \includegraphics[width=\linewidth]{graph_profile}
    %     \includegraphics[width=\linewidth]{graph_pressureLoss}
    %     \includegraphics[width=\linewidth]{graph_harvest}
    %     \end{center}
    %     \caption{
    %     Calculation of the power dissipation inside a typical water meter.
    %     The top figure shows flow profiles for one washing machine use, a shower and a toilet flush in a typical dwelling.
    %     Underneath is the head loss curve from a typical domestic water meter as a function of flow rate.
    %     The lower figure presents each of the water usage events (shower, washing machine and toilet flush) in terms of power dissipated inside the water meter due to mechanical losses.
    %     Dissipated power is calculated from the head loss curve and the flow rate data of the upper graphs.
    %     }
    %     \label{fig:profileSample}
    % \end{figure}

    \begin{figure}
        \begin{center}
        \includegraphics[width=\linewidth]{graph_profile}
        \end{center}
        \caption{Sample profile showing constructed instances of washing machine use, a shower and a toilet flush.
        Washing machine usage is broken into two parts corresponding to a wash and rinse cycle.}
        \label{fig:profileSample}
    \end{figure}

@@

    % \section{Estimation of Streaming Cell Harvester}
    % \label{sect:harvesterSize}

    \section{Conclusion}
    \label{sect:conclusion}
    We have calculated approximately \SI{1}{\kilo\joule} of energy is dissipated per day inside a typical water meter at a two occupant dwelling.
    We have estimated that a harvester would need to harvest approximately \SI{10}{\joule} of energy per day to record and transmit water consumption data.
    Therefore an energy harvester with an energy conversion efficiency of 1\% is required.
    Streaming cells we constructed produced conversion efficiencies in the order of $1\times 10^{-6}\%$.
